Assessing spatial correlations in pulsar timing arrays with the noise-marginalized optimal statistic

### 2017-08-04 manuscript/manuscript.tex
edits to text, especially Introduction

Low-frequency gravitational waves (GWs) with frequencies of 
$10^{-9} - 10^{-7} \; \mathrm{Hz}$ can be observed with pulsar timing arrays (PTAs) 
composed of millisecond pulsars (MSPs) \sv{citation}. 
The dominant astrophysical source in this frequency range is an isotropic stochastic 
gravitational wave background (GWB) 
made up of the incoherent superposition of GWs from inspiraling 
supermassive black hole binaries (SMBHBs) 
\citep{1995ApJ...446..543R, 2003ApJ...583..616J, 2003ApJ...590..691W}. 
By monitoring the ultra-regular periodic emission from these pulsars using radio telescopes, 
we can probe the dynamics of the spacetime through which the pulses travel. 
This is done by computing the pulsar timing residuals, 
which measure the difference between the expected and observed pulse times of arrival.

The primary data analysis technique used by PTAs is Bayesian inference 
to compare the probabilities of various models for the pulsar timing residuals, 
including one where the residuals contain evidence of an isotropic GW stochastic background. 
Bayesian inference is a powerful technique because it allows us to compare arbitrarily complicated 
models; however, they are computationally intensive, 
particularly when searching for evidence of the spatial correlations characteristic of a 
stochastic GWB -- ``the smoking gun'' of a stochastic GWB.

@@

When constructing the residuals, 
we typically fix the red noise parameters to the values that 
maximize the likelihood. 
However, this leads to a bias in the optimal statistic and SNR because 
the individual red noise parameters and the common red noise parameters 
are highly covariant. 
In this section, we compare three different techniques for computing the 
optimal statistic for PTAs with significant red noise. 
First, we fix the individual pulsars' red noise parameters to the 
maximum-likelihood values from individual pulsar analyses. 
Second, we fix the individual pulsars' red noise parameters to the 
values maximum-likelihood values from a common analysis 
involving all of the pulsars in our PTA. 
For the third method, we

In this paper we introduce a procedure to correct this bias by 
marginalizing over the red noise parameters using 
posterior distributions from a Bayesian analysis.

In order to compare the noise-marginalized optimal statistic to the 
fixed-noise optimal statistic, we injected a GW stochastic background with 
$\Agw = 5 \times 10^{-15}$ into simulated PTA data. 
We constructed simulated data sets using the same procedure as in \citet{tlb+2017}, 
which consisted of 10 MSPs with \sv{Steve fill in here...} 
The optimal statistic was computed using the \sv{fill in...} in PAL2 \sv{add citation}. 
We incorporate the individual pulsar noise parameters as follows. 
We perform individual Bayesian noise analyses on all of the pulsars, and 
fix the white noise to the maximum likelihood value. 
For the noise-marginalized optimal statistic, we draw the red noise parameters from 
the common red noise posteriors in order to average the optimal statistic over 
the individual pulsars' red noise. 
For comparison, we also compute the fixed-noise optimal statistic in two ways -- 
first, by fixing the red noise parameters at the maximum-likelihood values from individual pulsar analyses; 
second, by fixing the red noise parameters at the maximum-likelihood values from a common analysis.

>>>>

Low-frequency gravitational waves (GWs) with frequencies of 
$10^{-9} - 10^{-7} \; \mathrm{Hz}$ can be observed with pulsar timing arrays (PTAs) 
composed of millisecond pulsars (MSPs) \sv{citation}. 
The dominant astrophysical source in this frequency range is an isotropic stochastic 
gravitational wave background (GWB) 
made up of the incoherent superposition of GWs from inspiraling 
supermassive black hole binaries (SMBHBs) 
\citep{1995ApJ...446..543R, 2003ApJ...583..616J, 2003ApJ...590..691W}. 
By monitoring the ultra-regular periodic emission from these pulsars using radio telescopes, 
we can probe the dynamics of the spacetime through which the pulses travel. 
This is done by searching for correlations in the pulsar timing residuals, 
which measure the difference between the expected and observed pulse times of arrival.

The primary data analysis technique used by PTAs is Bayesian inference 
to compare the probabilities of various models for the pulsar timing residuals, 
including one where the residuals contain evidence of an isotropic GW stochastic background. 
Bayesian inference is a powerful technique because it allows us to compare arbitrarily complicated 
models; however, they are computationally intensive, 
particularly when searching for evidence of dipole spatial correlations -- 
``the smoking gun'' of a stochastic GWB.

@@

When constructing the residuals, 
we typically fix the red noise parameters to the values that 
maximize the likelihood. 
However, this leads to a bias in the optimal statistic and SNR because 
the individual red noise parameters and the common red noise parameters 
are highly covariant. 
In this section, we compare three different techniques for computing the 
optimal statistic for PTAs with significant red noise. 
First, we fix the individual pulsars' red noise parameters to the 
maximum-likelihood values from individual pulsar analyses. 
Second, we fix the individual pulsars' red noise parameters to the 
values maximum-likelihood values from a common analysis 
involving all of the pulsars in our PTA. 
For the third method, we draw values of the individual pulsars' 
red noise parameters from the posteriors generated by a Bayesian analysis 
that includes both individual red noise and a common red noise process, 
allowing us to marginalize over the individual red noise.

We used these three techniques to compute the optimal statistic for simulated data sets. 
We constructed simulated data sets using the same procedure as in \citet{tlb+2017}. 
The simulated PTAs consisted of 10 MSPs with \sv{Steve fill in here...}, 
and the injected GWB amplitude was $\Agw = 5\times10^{-15}$. 
The optimal statistic was computed using the \sv{fill in...} in PAL2 \sv{add citation}.
